Runtime-Assisted Convergence in Replicated Data Types

### 2022-03-15 evaluation.tex
Incorporating reviewer comments

We present an evaluation study of our runtime-assisted approach to
convergent RDTs with help of our prototype \quark implementation. The
benchmark we chose for the study is a collaborative document editing
application -- a common usecase addressed by several CRDT
proposals~\cite{rga, treedoc, crdts}. However, unlike these
other approaches, we did not have to build a dedicated replicated data
type to represent collaboratively-edited documents; an ordinary
document format extended with a merge operation would suffice. While
many data structures exist to represent text documents (e.g.,
ropes~\cite{boehm95}), we decided to adopt the simplest representation
of a document as a list of characters. 
\begin{center}
\begin{ocaml}
        type doc = char list
\end{ocaml}
\end{center}
While being simple, the advantage of this presentation is that we can
simply reuse the three-way \C{List.merge} function of the list data
type to merge documents. \C{List.merge} is a simple implementation of
the GNU \C{diff3} algorithm~\cite{gnudiff} in ~60 lines of OCaml. We
thus adopt a straightforward approach to building a collaborative
document editor with the intention to keep the development effort low
enough to be easily replicated. The convergence guarantee of \quark
ensures that the simplicity of our implementation doesn't come at the
expense of correctness. We now describe an evaluation that
demonstrates that it also doesn't come at the expense of the editor
performance.

@@

\quark's background merges however pose a new problem as they create
new versions on the local branch in the background while the user is
busy editing an older version. When the user attempts to write their
version of the document to the store, simply committing it would
effectively override the concurrent updates from other users obtained
via background merges. The solution, fortunately, is straightforward:
we \emph{merge} the user-submitted value with the (value of) the
latest version on the local branch to create a new version that
includes the updates from either direction. 
% The LCA for this merge is
% simply the last version on the local branch read by the user. This
% merge need not be synchronized as it doesn't alter the LCAs between
% any two branches. The merged value can then returned to the user as
% the result of the write.
Thus, \quark's \C{write} is a function of type $\texttt{Value}
\rightarrow \texttt{Value}$, where $\texttt{Value}$ is \C{doc} in the
current application.

\begin{figure}[ht]
  \centering
    \includegraphics[scale=0.3]{Figures/latency}
\caption{Latency of collaborative editing operations under \quark vs
  executing them with strong consistency (SC). }
\label{fig:latency}
\end{figure}

To measure the impact of \quark runtime on user writes, we measure the
latency of the \C{write} operation, which includes for the time spent
merging the user version with the current version, and persisting the
resultant version to the store. We conduct the experiments on a
three-node cluster of \C{i3.large} machines in Amazon \C{us-west2}
data center. Each user connects to one of the machines and performs
1000 edits in succession, saving the document after each edit. We
progressively increase the number of concurrent users editing the
document from 3 to 60 and measure the impact of the increased
concurrency on write latency. Fig.~\ref{fig:latency} shows the median
latency values. As evident from the figure, latency of \quark writes
remain more-or-less constant with the median latency around 0.008s.
The maximum latency ever measured for a \quark writes is 0.016s.
Notably, latency remains an order of magnitude less than the time
between consecutive edits (0.25s), making it hard to perceive.
Fig.~\ref{fig:latency} also plots the latency values for the baseline
``SC'' approach which achieves convergence by synchronizing each
operation, i.e., executing under strong consistency (SC). The SC write
latency increases super-linearly with the median value of 1.76s for 60
concurrent editors. The maximum latency measured for an SC write is
9.02s, which is considerably more than the inter-edit latency of
0.25s. The experiment confirms the theory that convergence does not
come at the cost of latency in
\quark.

\begin{figure}[ht]
  \centering
    \includegraphics[scale=0.3]{Figures/staleness}
\caption{Staleness increases as the number of concurrent editors
  increase.}
\label{fig:staleness}
  \vspace*{-0.2in}
\end{figure}

\quark system replicates the contents of each branch across all the
replicas as fast as the network allows. However, for a user $A$ to
see the changes made by the other user $B$, the changes have to be
reflected in $A$'s local version, which can only happen through a
merge operation. Since \quark synchronizes merge operations globally,
it induces additional delay before $A$ can see $B$'s changes. We call
this additional delay \emph{staleness} as with the progression of
time, $B$'s version known to $A$ becomes increasingly stale. At the
system-level, an increase in staleness effectively delays the
convergence (but doesn't preempt it, as proved by
Theorem~\ref{thm:progress}). To understand the staleness behavior in
\quark, we repeat the collaborative editing experiment, this time
measuring the staleness value at every merge. We do this by annotating
every version $v$ with the timestamp $t$ of its creation time. When
$v$ is is merged into a remote branch $b$ at a later time $t'$, the
difference $t' - t$ denotes the staleness of $v$ w.r.t the new version
on $b$. Multiple such staleness measurements are recorded for each
experiment to compute the 10th, 50th, and 90th percentile
values. Fig.~\ref{fig:staleness} shows the results. While
staleness remains in the order of milliseconds with fewer ($\leq 9$)
concurrent users, it increases super-linearly as we increase the
number of collaborators in the multiples of 3 until 30. While
increased staleness is inevitable in our approach due to synchronized
merges, we believe the increase can be contained by choosing the order
of merges to avoid the ``starvation'' of some branches. Note that
relaxing the linearizability constraint on merges would completely
eliminate the staleness overhead, but the resultant Git-like system
fails to converge due to anomalous executions described in
Sec.~\ref{sec:motivation}\footnote{
  Git admits anamolous version history graphs where two branches can
  have the same set of commits and yet differ in their final version.
  Supplementary material describes two such cases we observed on
  Github.}.

>>>>

In this section we present an emperical evaluation of \quark on two
case studies. First is a collaborative document editing
application -- a common usecase addressed by several CRDT
proposals~\cite{rga, treedoc, crdts}. Second is a replicated key-value
store implemented using a mergeable Red-Black tree data structure.

\subsection{Collaborative Editing}

\quark's MRDT approach obviates the need to build a dedicated
replicated data type for collaboratively-edited documents; an ordinary
document format extended with a merge operation would suffice. While
many data structures exist to represent text documents (e.g.,
ropes~\cite{boehm95}), we decided to adopt the simplest representation
of a document as a list of characters. 
\begin{center}
\begin{ocaml}
        type doc = char list
\end{ocaml}
\end{center}
While being simple, the advantage of this presentation is that we can
simply reuse the three-way \C{List.merge} function of the list data
type to merge documents. \C{List.merge} is a simple implementation of
list merge algorithm (in ~60 lines of OCaml) inspired by the GNU
\C{diff3} algorithm~\cite{gnudiff}. We thus adopt a straightforward
approach to building a collaborative document editor with the
intention to keep the development effort low enough to be easily
replicated. The convergence guarantee of \quark ensures that the
simplicity of our implementation doesn't come at the expense of
correctness. The aim of the experimental evaluation is to quantify the
impact of \quark on the performance.

@@

\quark's background merges however pose a new problem as they create
new versions on the local branch in the background while the user is
busy editing an older version. When the user attempts to write their
version of the document to the store, simply committing it would
effectively override the concurrent updates from other users obtained
via background merges. The solution, fortunately, is straightforward:
we \emph{merge} the user-submitted value with the (value of) the
latest version on the local branch to create a new version that
includes the updates from either direction. Since this merge is fully
confined to the local replica, it is guaranteed to not affect the LCA
of the local branch with respect to a remote branch. To the external
world, it appears as if the local branch has simply committed a new
version.
% The LCA for this merge is
% simply the last version on the local branch read by the user. This
% merge need not be synchronized as it doesn't alter the LCAs between
% any two branches. The merged value can then returned to the user as
% the result of the write.
Thus, \quark's \C{write} is a function of type $\texttt{Value}
\rightarrow \texttt{Value}$, where $\texttt{Value}$ is \C{doc} in the
current application.

\noindent\paragraph{Latency} To measure the impact of \quark runtime on user
writes, we measure the latency of the \C{write} operation, which
includes the time spent merging the user version with the current
version, and persisting the resultant version to the store. We conduct
the experiments on a three-node cluster of \C{i3.large} machines in
Amazon \C{us-west2} data center. Each user connects to one of the
machines, forks a new branch, and performs 1000 edits in succession,
saving the document after each edit. We progressively increase the
number of concurrent users editing the document from 3 to 200 and
measure the impact of the increased concurrency on write latency.
Fig.~\ref{fig:latency} shows the 10th, 50th (median), and 90th
percentile latency values. The median and 10th percentile latencies
remain more-or-less constant with a slight increase between 3 and 60
concurrent users. This demonstrates that \quark has negligible effect
on the per-operation latency, which is expected. The 90th percentile
latency, however, shows no clear pattern by the virtue of being more
susceptible to transient system conditions. Nonetheless, the maximum
value of 90th percentile latency measured (32ms) remains well below
the the time between consecutive edits (0.25s), making it hard to
perceive by a human user.

\begin{figure}[ht]
  \centering
    \includegraphics[scale=0.4]{Figures/monkey_latency}
\caption{Latency of writes to a shared document under \quark. }
\label{fig:latency}
\end{figure}

For comparison against a baseline, we have implemented an ``SC''
approach which achieves convergence by synchronizing each operation,
i.e., executing it under strong consistency (SC). The SC
implementation shares most of its code with \quark with the only
change being that it wraps commits instead of merges inside a lock.
As with \quark, we measured the write latency of the SC implementation
while increasing the number of concurrent users ($n$).  The median SC
write latency increases super-linearly from 10ms for $n=3$ to 63s
(i.e., >1m) for $n=60$, which is considerably more than the inter-edit
latency of 0.25s.

\noindent\paragraph{Staleness} Our implementation of \quark relies on
Scylla to replicate the contents of each branch across all the
replicas as fast as the network allows. However, for a user $A$ to see
the changes made by the other user $B$, the changes have to be
reflected in $A$'s local version, which can only happen through a
merge operation. Since \quark synchronizes merge operations globally,
it induces additional delay before $A$ can see $B$'s changes. We call
this additional delay \emph{staleness} as with the progression of
time, $B$'s version known to $A$ becomes increasingly stale. At the
system-level, an increase in staleness effectively delays the
convergence (but doesn't preempt it, as proved by
Theorem~\ref{thm:progress}).

\begin{figure}[ht]
  \centering
    \includegraphics[scale=0.4]{Figures/monkey_staleness}
\caption{Staleness increases as the number of concurrent editors
  increase.}
\label{fig:monkey-staleness}
  \vspace*{-0.2in}
\end{figure}

To understand the effect of \quark on
staleness, we quantify and measure it along with latency in the
experiment setup described above. Staleness is defined as the time
taken for a version committed on one replica to be merged into a
concurrent version on a remote replica. To measure staleness, we
annotate every version $v$ with the timestamp $t$ of capturing the
wall clock time of its creation. When $v$ is is merged into a remote
branch $b$ at a later time $t'$, the difference $t' - t$ denotes the
staleness of $v$ w.r.t the new version on $b$. One such staleness
measurement is recorded for every merge that ever happens during the
experiment. We compute 10th, 50th, and 90th percentiles of staleness
values thus obtained. Fig.~\ref{fig:monkey-staleness} shows the
results. As evident, staleness increases steadily with the increasing
number of concurrent replicas, which is expected considering that
merges are synchronized, and increasing the number of replicas reduces
the number of opportunities for a replica to merge. The increase is
roughly linear in the number of replicas as our implementation passes
the lock around in a round-robin fashion, thus increasing the lock
latency in proportion to the number of replicas. Nonetheless, the 90th
percentile staleness values remain low -- in the order of 100ms with
the number of replicas under 30, and in the order of 1s with number of
replicas under 60. While further optimizations might reduce staleness,
a non-trivial staleness overhead is inevitable in \quark due to our
use of synchronized merges to guarantee convergence.

% Our approach thus
% introduces staleness as novel tradeoff against convergence. While
% trading off staleness may not be an optimal choice for every
% application, it is arguably a less disruptive choice if an application
% has to choose between latency and staleness.
% Sec.~\ref{sec:motivation}\footnote{
%   Git admits anamolous version history graphs where two branches can
%   have the same set of commits and yet differ in their final version.
%   Supplementary material describes two such cases we observed on
%   Github.}.

\subsection{Key-Value Store}

We repeat our experiments on a mergeable key-value store application
we implemented using a Red-Black tree MRDT. Latency results are shown
in Fig.~\ref{fig:rb-latency}, and staleness results in
Fig.~\ref{fig:rb-staleness}.

\begin{figure}[ht]
  \centering
    \includegraphics[scale=0.4]{Figures/rbmonkey_latency}
  \caption{Latency for Key-Value store operations under \quark}
\label{fig:rb-latency}
  \vspace*{-0.2in}
\end{figure}

\begin{figure}[ht]
  \centering
    \includegraphics[scale=0.4]{Figures/rbmonkey_staleness}
  \caption{Key-Value store staleness under \quark}
\label{fig:rb-staleness}
  \vspace*{-0.2in}
\end{figure}

### 2022-03-15 implementation.tex
Incorporating reviewer comments

We realize a prototype of \quark runtime as a lightweight shim layer
on top of Scylla -- an off-the-shelf distributed data
store~\cite{scylla}. We rely on Scylla for inter-replica
communication, data replication, persistence, and fault tolerance.
\quark translates the high-level mergeable data type implementations
in OCaml to their low-level representations in the backing store and
organizes their well-formed distributed executions.
Fig.~\ref{fig:implementation} illustrates the overall architecture.

The implementation of \quark the is a straightforward realization of
the \quark distributed machine described in Sec.~\ref{sec:concrete-sem}.
We manifest each component of the state, namely the branch map $B$,
the value map $N$, and the head map $H$, as a column family (i.e., a
table) in Scylla. The synchronization needed to linearize merges is
implemented with help of Scylla's support for conditional updates (CAS
operations) and expiring columns. The total order among merges is
enforced with help of \C{Quorum} reads and writes. Each user process
is assigned its own replica of the RDT.  Version vectors are realized
as associative lists and stored in Scylla
as blobs.

One key difference between the formalization and the implementation is
in the treatment of RDT values. Formalization assumes values to be
atomic with no sharing in between them. In practice, however, an RDT
value could be a linked data structure such as a binary tree, and two
such values could share a significant amount of internal structure.
Consequently, the size of the \emph{diff} between two consecutive
versions of a value could be asymptotically less then the size of the
data structure itself, in which case it unreasonable to transfer the
entire data structure over the network. To facilitate the efficient
computation of diffs between versions of data structures, we implement
a \emph{content-addressable store} as a key-value table in Scylla
where key is simply the SHA256 hash of the value. A linked data
structure is stored as a collection of nodes, where each node links to
the other by referring to its hash. The diff between two consecutive
versions of a data structure would simply manifest as new entries in
the content-addressable store reachable from the root of the new
version. The new entries, being new data, are automatically replicated
by Scylla, thus letting us reconstruct the new version at a remote
location. The root hash of the new version is obtained from the value
map $N$, which now maps version vectors to the \emph{hashes} of
values.

>>>>

We realize a prototype of \quark runtime for MRDTs as a lightweight
shim layer on top of Scylla -- an off-the-shelf distributed data
store~\cite{scylla}. We rely on Scylla for inter-replica
communication, data replication, persistence, and fault tolerance.
\quark translates the high-level MRDT implementations in OCaml to
their low-level representations in the backing store and orchestrates
their well-formed distributed executions.
Fig.~\ref{fig:implementation} illustrates the overall architecture.

The implementation of \quark is a straightforward realization of the
\quark distributed machine described in Sec.~\ref{sec:concrete-sem}.
We manifest each component of the state, namely the branch map $B$,
the value map $N$, and the head map $H$, as a column family (i.e., a
table) in Scylla. The synchronization needed to linearize merges is
implemented with help of Scylla's support for conditional updates (CAS
operations) and expiring columns. The total order among merges is
enforced with help of \C{Quorum} reads and writes. Each user process
is assigned its own replica of the MRDT.  Version vectors are realized
as associative lists and stored in Scylla as blobs.

One key difference between the formalization and the implementation is
in the treatment of MRDT values. Formalization assumes values to be
atomic with no sharing in between them. In practice, however, an MRDT
could be a linked data structure such as a binary tree, and two
such values could share a significant amount of internal structure.
Consequently, the size of the \emph{diff} between two consecutive
versions of a value could be asymptotically less then the size of the
data structure itself, in which case it unreasonable to transfer the
entire data structure over the network. To facilitate the efficient
computation of diffs between versions of data structures, we implement
a \emph{content-addressable store} as a key-value table in Scylla
where key is simply the SHA256 hash of the value. A linked data
structure is stored as a collection of nodes, where each node links to
the other by referring to its hash. The diff between two consecutive
versions of a data structure would simply manifest as new entries in
the content-addressable store reachable from the root of the new
version. The new entries, being new data, are automatically replicated
by Scylla, thus letting us reconstruct the new version at a remote
location. The root hash of the new version is obtained from the value
map $N$, which now maps version vectors to the \emph{hashes} of
values.

### 2022-03-15 introduction.tex
Committing last minute edits.
Incorporating reviewer comments

% hinders the widespread adoption of replicated data
% types in software development.
In this paper we propose such an alternative approach to convergent
replicated data types. The key distinguishing characteristic of our
approach is the utilization of a distributed runtime to orchestrate
\emph{well-formed} distributed executions that are guaranteed to
converge. We define well-formedness as a constraint over the
\emph{states} that arise during the execution as opposed to the
\emph{operations} that are involved in the execution. To allow for its
enforcement, we adopt a state-centric model of replication based on
version-controlled \emph{mergeable} states~\cite{mrdt} instead of an
operation-centric model based on commutative operations. Unlike
commutativity, mergeability does not require a data type definition to
be refactored to suit distributed execution. Instead, the type
definition only needs to be extended with a \C{merge} function to
reconcile concurrent versions of the type in presence of their common
ancestor version. Notably, \C{merge} function is left completely
unconstrained. Consequently, programmer can chose any merge semantics
that make sense within the context of the data type's (sequential)
semantics without risking the loss of convergence. Such strong
guarantees are made possible by the simplicity of state-based
reasoning combined with the run-time enforcement of well-formedness.

% combined with the simplicity
% of state-based reasoning. Note that state-centric model of
% replication has also been explored in the context of CRDTs. However,
% state-based CRDTs require the replicated state to be organized as a
% lattice -- a restriction we relax in our setting.

Note that it is easy to construct a trivial runtime that guarantees
convergence of all executions.  Such a runtime would make an extensive
use of inter-replica coordination to synchronize the execution of
every operation and thus induce a sequentially-consistent behavior.
This would however defeat the purpose of state replication as the
latency incurred for synchronized execution of an operation is
prohibitively high, and the availability of the system during the
execution is correspondingly low. It is therefore important for a
distributed runtime of RDTs to not synchronize the execution of RDT
operations. Orchestrating a well-formed convergent execution without
impacting latency and availability may seem impossible due to the CAP
theorem~\cite{cap}, but it is not the case in fact. While CAP theorem
does rule out linearizability, it does not prohibit enforcing weaker
consistency constraints on distributed executions. Our approach
exploits this observation by (\rom{1}). Identifying a novel set of
constraints on distributed executions that guarantee their
convergence, and (\rom{2}). Localizing such constraints on
(asynchronously executed) state merges such that per-operation latency
and system-wide availability remain unaffected.
Sec.~\ref{sec:concrete-sem} explains our runtime in detail.

% In fact many CRDTs,
% including the aforementioned \emph{remove-wins} set, implicitly assume
% \emph{causal consistency}, which is weaker than linearizability but is
% stronger than the default \emph{eventual consistency}. While causal
% consistency alone cannot guarantee convergence, it sufficiently
% constrains the distributed execution for the CRDT implementations to
% guarantee convergence. In this respect our approach differs in the
% sense that

\paragraph{Contributions} The contributions of this paper are summarized
below:
\begin{itemize}
  \item We present a runtime-assisted approach to state replication
    that sufficiently constrains distributed executions so as to
    guarantee the convergence of replicated data types. This is in
    contrast to the existing models of replication, where each type is
    obligated to prove its convergence.

  \item We formalize the aforementioned constraints in the context of
    a an abstract machine (named \quark) that manifests distributed
    executions as version history graphs (similar to Git).
    We prove the \emph{convergence} and \emph{progress} properties of
    the \quark abstract machine that together guarantee its soundness.
    The formalization has been mechanized in Ivy~\cite{ivy} and
    \emph{automatically} verified with help of Z3~\cite{z3}. To the
    best of our knowledge, this is the first time an SMT solver was
    used to verify the \emph{meta-theory} of a formal system.

@@

  \item We describe an implementation of \quark as a light-weight shim
    layer on top of Scylla -- an off-the-shelf distributed key-value
    store~\cite{scylla}. We evaluate the implementation on a
    collaborative-editing benchmark to demonstrate novel trade-offs
    presented by our approach.
\end{itemize}

>>>>

% In this paper we propose such an alternative approach to convergent
% replicated data types. The key distinguishing characteristic of our
% approach is the utilization of a distributed runtime to orchestrate
% \emph{well-formed} distributed executions that are guaranteed to
% converge. We define well-formedness as a constraint over the
% \emph{states} that arise during the execution as opposed to the
% \emph{operations} that are involved in the execution. To allow for its
% enforcement, we adopt a state-centric model of replication based on
% version-controlled \emph{mergeable} states~\cite{mrdt} instead of an
% operation-centric model based on commutative operations. Unlike
% commutativity, mergeability does not require a data type definition to
% be refactored to suit distributed execution. 
In this paper we describe an alternative approach to building
convergent replicated data types that realizes the aforementioned
virtues. Our approach is based on Mergeable Replicated Data Types
(MRDTs)~\cite{mrdt} -- an alternative take on RDTs that is inspired by
the Git version control system. MRDTs adopt a state-centric model of
replication based on version-controlled \emph{mergeable} states
instead of an operation-centric model based on commutative operations.
Unlike commutativity, mergeability does not require a data type
definition to be refactored to suit distributed execution. Instead,
the type definition only needs to be extended with a \C{merge}
function to reconcile concurrent versions of the type in presence of
their common ancestor version. However, mergeability itself doesn't
guarantee convergence; there exist MRDTs with well-defined and
intuitive merge semantics that nonetheless admit divergent executions.
While constraining the data type semantics to conform to algebraic
laws such as monotonicity might restore convergence, it would lead
MRDTs down the path of CRDTs, which has known problems described
above. We therefore propose an alternative approach to convergence --
one that is based on runtime orchestration rather than static
enforcement. In particular, we extend MRDTs with a distributed runtime
that orchestrates only the \emph{well-formed} executions which are
guaranteed to converge. Notably, our approach guarantees convergence
of MRDTs \emph{regardless} of their merge semantics, thus obviating
the need to constrain the type semantics or their implementations. The
key insight behind our approach is the observation that a data type's
\C{merge} is a complete function, and thefore can be safely invoked to
merge concurrent versions of the type notwithstanding their individual
operation histories. Thus, a replica's current version
can always be made available for user operations with the guarantee
that it can be safely merged with a concurrent version at a later
point of time. The ability to arbitrarily delay merges gives us the
opportunity to orchestrate them in a way that is \emph{effectively}
linear, which inturn guarantees convergence.
% combined with the simplicity
% of state-based reasoning. Note that state-centric model of
% replication has also been explored in the context of CRDTs. However,
% state-based CRDTs require the replicated state to be organized as a
% lattice -- a restriction we relax in our setting.

Note that it is easy to construct a trivial runtime that guarantees
convergence of all executions. Such a runtime would make an extensive
use of inter-replica coordination to synchronize the execution of
every operation and thus induce a sequentially-consistent behavior.
This would however defeat the purpose of state replication as the
latency incurred for synchronized execution of an operation is
prohibitively high, and the availability of the system during the
execution is correspondingly low. It is therefore important for a
distributed runtime of RDTs to \emph{not} interfere with the execution
of RDT operations. Orchestrating a well-formed convergent execution
without impacting latency and availability may seem impossible due to
the CAP theorem~\cite{cap}, but it is not the case in fact. While CAP
theorem does rule out linearizability, it does not prohibit enforcing
weaker consistency constraints on distributed executions. Our approach
exploits this observation by (\rom{1}).  Identifying a novel set of
constraints on distributed executions that guarantee their
convergence, and (\rom{2}). Localizing such constraints on
(asynchronously executed) state merges such that per-operation latency
and system-wide availability remain unaffected. The downside of
interfering with merges, however, is that the replicas may now take
longer to converge -- a phenomenon we quantify using a metric called
\emph{staleness} (Sec.~\ref{sec:eval}). Our approach thus brings to
fore a fundamental tradeoff that was never explored in the context of
RDTs before.

%Sec.~\ref{sec:motivation} explains these insights in detail.

\paragraph{Contributions} The contributions of this paper are summarized
below:
\begin{itemize}
  \item We present a runtime-assisted approach to MRDT state
    replication that sufficiently constrains distributed executions so
    as to guarantee their convergence. This is in contrast to the
    existing models of replication, where each data type is obligated
    to prove its convergence by discharging algebraic proof
    obligations.

  \item We formalize the aforementioned constraints in the context of
    an abstract machine (named \quark) that manifests distributed
    executions as version history graphs similar to Git.  We prove the
    \emph{convergence} and \emph{progress} properties of the \quark
    abstract machine that together guarantee its soundness. The
    formalization has also been mechanized in Ivy~\cite{ivy} and
    automatically verified with help of Z3~\cite{z3}. To the best of
    our knowledge, this is the first time the meta-theory of a formal
    is proven completely inside an SMT solver.

@@

  \item We describe an implementation of \quark as a light-weight shim
    layer on top of Scylla -- an off-the-shelf distributed key-value
    store~\cite{scylla}. We describe an extensive evaluation to
    quantify the staleness metric and novel tradeoffs in the context
    of case studies involving a data structure benchmark and a
    collaborative editing application. 
\end{itemize}
